---
title: "DEG on Type 1 Diabetes Associated Bacteria"
author: "Pavalya"
date: "2025-05-07"
output: html_document
---

```{r}
#load libraries
library(DESeq2)
library(ggplot2)
library(GEOquery)
```
# Step 1: Preparing count data
```{r}
#read in counts data
counts <- read.delim("type1diabetes_rawcounts.txt", header=TRUE, row.names=1, sep="\t")
head(counts)
# Load the count file (if not already loaded)
counts <- read.delim("type1diabetes_rawcounts.txt", header = TRUE, row.names = 1, sep = "\t", check.names = FALSE)

# Remove ".1", ".2", etc. from column names
cleaned_colnames <- sub("\\.\\d+$", "", colnames(counts))

# View unique sample names
unique(cleaned_colnames)

#Collapse technical replicates by summing columns
# Replace colnames with cleaned (replicate-free) names
colnames(counts) <- cleaned_colnames
# Collapse duplicate columns by summing them
collapsed_counts <- as.data.frame(t(rowsum(t(counts), group = colnames(counts))))
#check result
dim(collapsed_counts)
head(colnames(collapsed_counts))



```

```{r}
#read in sample info
# Obtain metadata
library(tidyverse)
sample_info <- read_csv("SraRunTable.csv", show_col_types = FALSE)
View(sample_info)
```

```{r}
# Manipulate metadata

library(dplyr)

 metadata_sub <- sample_info %>%
  dplyr::select(
    agent,
    'GEO_Accession (exp)',
    source_name,
    time_point)
    
 
View(metadata_sub)
# Step 3.1: Rename GEO accession column for ease of use


metadata_sub <- metadata_sub %>%
  rename_with(~ "accession", .cols = matches("GEO_Accession"))



#Step 4: Match metadata to the collapsed count matrix
# Keep only metadata entries that match the count column names
metadata_filtered <- metadata_sub %>%
  filter(accession %in% colnames(collapsed_counts))

# Reorder rows of metadata to match column order in count matrix
metadata_filtered <- metadata_filtered[match(colnames(collapsed_counts), metadata_filtered$accession), ]

# Check alignment
all(metadata_filtered$accession == colnames(collapsed_counts))  # should return TRUE

# making sure the row names in metadata matches to the columns names in counts data 
all(metadata_filtered$accession %in% colnames(collapsed_counts))  # should return TRUE



```
```{r}
library(DESeq2)
```

#Step 5: Create DESeq2 dataset
```{r}
# Create DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = collapsed_counts,
                              colData = metadata_filtered,
                              design = ~ agent)  # or ~ agent + time_point if you want both

# Clean up agent levels: remove parentheses, replace spaces with underscores
dds$agent <- gsub("\\s+", "_", dds$agent)      # Replace spaces with underscores
dds$agent <- gsub("[()]", "", dds$agent)       # Remove parentheses
dds$agent <- factor(dds$agent)                 # Re-factor the cleaned labels

# Check result
levels(dds$agent)

dds

```

# prefiltering: removing rows with low gene counts
#keeping rows that has 10 reads total 
```{r}
keep <- rowSums(counts(dds)) >=10 # True or false(logical values)

# use the logical values to subset deseq dataset object
dds <- dds[keep,]

dds

```
# set the factor level
```{r}
#To manually set the reference level:
dds$agent <- relevel(dds$agent, ref = "Serum-Free_Control")
dds$agent

```

# run DESeq
```{r}
dds <- DESeq(dds)
```
#To Normalize raw counts for visualization:
```{r}
vsdata <- vst(dds, blind=FALSE)
```
# Lets again look at the qc of our data 
#First lets do PCA plot
```{r}
plotPCA(vsdata,intgroup="agent" )
plotPCA(vsdata,intgroup="time_point" )


#improve
# Load required libraries
library(DESeq2)
library(ggplot2)

# # STEP 1: Variance Stabilizing Transformation (already done)
# vsdata <- vst(dds, blind = FALSE)
# 
# # STEP 2: Select top 1000 most variable genes (optional but improves clarity)
# top_var_genes <- head(order(rowVars(assay(vsdata)), decreasing = TRUE), 1000)
# vsdata_top <- vsdata[top_var_genes, ]
# 
# # STEP 3: Get PCA data with sample metadata
# pcaData <- plotPCA(vsdata_top, intgroup = "agent", returnData = TRUE)
# percentVar <- round(100 * attr(pcaData, "percentVar"))
# 
# # STEP 4: Custom PCA plot using ggplot2
# ggplot(pcaData, aes(x = PC1, y = PC2, color = agent)) +
#   geom_point(size = 3) +
#   xlab(paste0("PC1: ", percentVar[1], "% variance")) +
#   ylab(paste0("PC2: ", percentVar[2], "% variance")) +
#   ggtitle("PCA of Top 1000 Most Variable Genes") +
#   theme_bw() +
#   theme(
#     legend.position = "right",
#     plot.title = element_text(hjust = 0.5),
#     text = element_text(size = 12)
#   )
# 
# 
# 
# 
# pcaData <- plotPCA(vsdata_top, intgroup = c("agent", "time_point"), returnData = TRUE)
# 
# ggplot(pcaData, aes(x = PC1, y = PC2, color = agent)) +
#   geom_point(size = 3) +
#   facet_wrap(~ time_point) +
#   theme_bw()
# 
# 
# 


# ggplot(pcaData, aes(x = PC1, y = PC2, color = agent)) +
#   geom_point(size = 3, alpha = 0.9) +
#   facet_wrap(~ time_point) +
#   xlab(paste0("PC1: ", percentVar[1], "% variance")) +
#   ylab(paste0("PC2: ", percentVar[2], "% variance")) +
#   ggtitle("PCA by Treatment Agent and Time Point") +
#   theme_bw() +
#   theme(
#     plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
#     strip.text = element_text(size = 12),
#     axis.text = element_text(size = 10),
#     axis.title = element_text(size = 12),
#     legend.title = element_text(size = 12),
#     legend.text = element_text(size = 10)
#   )


#Make sure vsdata is defined:
vsdata <- vst(dds, blind = FALSE)
# Save output to PNG
png("PCAplot_agents_time.png", width = 1200, height = 1000, res = 200)
# Example with shape for agent and color for time_point
pcaData <- plotPCA(vsdata, intgroup = c("agent", "time_point"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData, aes(PC1, PC2, color = time_point, shape = agent)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA: Combined View by Time and Agent") +
  theme_bw()
dev.off()



```
#Volcano Plot: IL-1B vs Serum-Free_Control
```{r}
# # Run DESeq2 if not already done
# dds <- DESeq(dds)
# 
# # Set reference level
# dds$agent <- relevel(dds$agent, ref = "Serum-Free_Control")
#Make sure vsdata is defined:
vsdata <- vst(dds, blind = FALSE)
# Save output to PNG
png("volcanoplot_IL-1B_Serum-Free_Control.png", width = 1200, height = 1000, res = 300)

# Extract results: IL-1B vs Serum-Free_Control
res_IL1B <- results(dds, contrast = c("agent", "IL-1B", "Serum-Free_Control"))

# Remove NA and order by adjusted p-value
res_IL1B <- na.omit(res_IL1B)
res_IL1B$gene <- rownames(res_IL1B)
res_IL1B <- as.data.frame(res_IL1B)  # 🔧 Convert to plain data frame


# Volcano plot
library(ggplot2)
library(dplyr)

res_IL1B <- res_IL1B %>%
  mutate(
    sig = ifelse(padj < 0.05 & abs(log2FoldChange) > 1,
                 ifelse(log2FoldChange > 1, "Up", "Down"),
                 "NS")
  )


ggplot(res_IL1B, aes(x = log2FoldChange, y = -log10(padj), color = sig)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("Up" = "red", "Down" = "blue", "NS" = "grey")) +
  labs(title = "Volcano Plot: IL-1B vs Control",
       x = "log2 Fold Change",
       y = "-log10 Adjusted P-value") +
  theme_minimal()
dev.off()

```
#Volcano Plot: Bacteroides_dorei_BD vs Serum-Free_Control
```{r}
#Make sure vsdata is defined:
vsdata <- vst(dds, blind = FALSE)
# Save output to PNG
png("volcanoplot_Bacteroides_dorei_BD_Serum-Free_Control.png", width = 1200, height = 1000, res = 300)
# Extract results: Bacteroides dorei vs Control
res_BD <- results(dds, contrast = c("agent", "Bacteroides_dorei_BD", "Serum-Free_Control"))

# Remove NA values
res_BD <- na.omit(res_BD)

# Add gene IDs
res_BD$gene <- rownames(res_BD)

# Mark significance based on adjusted p-value and log2 fold change
res_BD_df <- as.data.frame(res_BD) %>%
  dplyr::mutate(sig = ifelse(padj < 0.05 & abs(log2FoldChange) > 1,
                             ifelse(log2FoldChange > 1, "Up", "Down"),
                             "NS"))

# Volcano plot
library(ggplot2)
ggplot(res_BD_df, aes(x = log2FoldChange, y = -log10(padj), color = sig)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("Up" = "red", "Down" = "blue", "NS" = "grey")) +
  labs(title = "Volcano Plot: Bacteroides dorei vs Control",
       x = "log2 Fold Change", y = "-log10 Adjusted P-value") +
  theme_minimal()
dev.off()

```
#Biological Insight:
This supports the hypothesis that Bacteroides dorei, a bacterium enriched in pre-diabetic gut microbiomes, triggers strong pro-inflammatory or immune-related responses in islets.

These DEGs could include cytokines, chemokines, or other signaling molecules relevant to autoimmunity.

#Heatmap of Top DEGs for All Bacterial Agents vs Control
```{r}
#Make sure vsdata is defined:
vsdata <- vst(dds, blind = FALSE)

# Load required libraries
library(DESeq2)
library(pheatmap)
library(dplyr)

# Save output to PNG
png("Heatmap_Top_DEGs_Across_Agents.png", width = 1200, height = 1000, res = 150)
vsdata <- vst(dds, blind = FALSE)



# Define bacterial agents and control
bacterial_agents <- c("Bacteroides_dorei_BD", "Escherichia_coli_EC", "Ruminococcus_gnavus_RG")
reference <- "Serum-Free_Control"
top_genes_all <- c()

# Extract top genes from each comparison
for (agent in bacterial_agents) {
  res <- results(dds, contrast = c("agent", agent, reference))
  res <- na.omit(res)

  sig_res <- res %>%
    as.data.frame() %>%
    filter(padj < 0.05 & abs(log2FoldChange) > 1) %>%
    arrange(padj)

  top_genes <- rownames(sig_res)[1:min(30, nrow(sig_res))]
  top_genes_all <- unique(c(top_genes_all, top_genes))
}

# Subset VST-normalized expression for top genes
vsdata_sub <- assay(vsdata)[top_genes_all, ]

# Create annotation dataframe
annotation_col <- as.data.frame(colData(vsdata)[, c("agent", "time_point")])

# Plot the heatmap
pheatmap(vsdata_sub,
         annotation_col = annotation_col,
         scale = "row",
         fontsize_row = 6,
         fontsize_col = 8,
         main = "Top DEGs across Bacterial Agents vs Control",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         show_rownames = FALSE)

dev.off()
```


# explore results
#save the results
```{r}
res <- results(dds)
# view the results
res
```
# logfold2 change calculated in desired factor agent with refernce as serum free control
# All the values here is with respect to Ruminococcus gnavus
#The statistical test used here is wald test 
# base mean: Average of the normalized counts taken over of all the samples
#log2 fold change is the change of the respective gene in the Ruminococcus gnavus comapred to Serum.Free Control.
#The positive values are upregulated genes in the Ruminococcus gnavus condition and negative values are downregulated genes in the Ruminococcus gnavus condition.
#ifcSE: Provides the standard error estimates of the log2fold change
# The stat here are the wald test values for these genes.
#p value is for the test statistic for this gene.
#padj value are corrected pvalue for multiple testing.
#Note: We need to correct the pvalue for multiple testing is because whenever we perform a statsistical test we use a p-value of 0.05, 5% of our DEG's are not really differentially expressed but they are only due to random chance and there is no real effect.
#In our dataset we have around 32839 genes, 5% of 32839 is  1,642 genes, these in our list of DEG's genes these  1,642 genes are false positives.
#To avoid detection of false positives ,these adjp values help
```{r}
#Summary for the result
summary(res)

```

```{r}
#change the padj values
res0.01 <- results(dds, alpha =0.01)
summary(res0.01)
```

#contrasts (more than 2 levels in a factor)
```{r}
results(dds, contrast=c("agent", "Bacteroides_dorei_BD", "Serum-Free_Control"))
```

# MA plot

```{r}
plotMA(res)
```
# MA plot is scatter plot of logfold change versus mean of normalized counts, shows us the genes that are differentially expressed.
# Blue colored ones are significantly differentially expressed, they have padj value of less than 0.05.
# Smaller traingles towards the edge of the plot , these genes have higher fold changes.
#Direction of the triangle tells us the direction of the fold change.
#genes on the upper and lower right quadrant means these genes have high mean of normalized counts and high log fold changes, make this as interesting candidates  to further loo into.

#Functional annoatation
```{r}
# Install if not already installed
# BiocManager::install("clusterProfiler")
# BiocManager::install("org.Hs.eg.db")

library(clusterProfiler)
library(org.Hs.eg.db)
library(dplyr)
library(enrichplot)

# Step 1: Extract significant gene IDs from IL-1B vs Control results
sig_genes <- res_IL1B %>%
  as.data.frame() %>%
  filter(padj < 0.05 & abs(log2FoldChange) > 1) %>%
  pull(gene)  # assumes gene column has Ensembl IDs

# Step 2: Convert Ensembl to Entrez IDs
gene_ids <- bitr(sig_genes,
                 fromType = "ENSEMBL",
                 toType = "ENTREZID",
                 OrgDb = org.Hs.eg.db)

# Check how many mapped successfully
cat("Mapped genes:", nrow(gene_ids), "out of", length(sig_genes), "\n")

# Step 3: Keep only successfully mapped Entrez IDs
mapped_genes <- unique(gene_ids$ENTREZID)

# Step 4: GO enrichment analysis (Biological Process)
ego <- enrichGO(gene         = mapped_genes,
                OrgDb        = org.Hs.eg.db,
                keyType      = "ENTREZID",
                ont          = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                readable      = TRUE)

# Step 5: KEGG pathway enrichment
ekegg <- enrichKEGG(gene         = mapped_genes,
                    organism     = 'hsa',
                    pvalueCutoff = 0.05)

# Step 6: Plot GO enrichment
barplot(ego, showCategory = 10, title = "GO: Biological Process")

# Step 7: Plot KEGG enrichment
barplot(ekegg, showCategory = 10, title = "KEGG Pathway Enrichment")


```
# Multi-Agent Enrichment Analysis
```{r}
# Load required packages
library(clusterProfiler)
library(org.Hs.eg.db)
library(dplyr)
library(DESeq2)
library(enrichplot)

# Step 1: Define your agents and reference
agents <- c("Bacteroides_dorei_BD", "Escherichia_coli_EC", "Ruminococcus_gnavus_RG")
reference <- "Serum-Free_Control"

# Step 2: Collect significant DEGs from all agents
all_sig_genes <- c()

for (agent in agents) {
  res <- results(dds, contrast = c("agent", agent, reference))
  res <- na.omit(res)
  
  sig_res <- res %>%
    as.data.frame() %>%
    filter(padj < 0.05 & abs(log2FoldChange) > 1)
  
  sig_res$gene <- rownames(sig_res)
  all_sig_genes <- c(all_sig_genes, sig_res$gene)
}

# Remove duplicates
all_sig_genes <- unique(all_sig_genes)

# Step 3: Convert Ensembl IDs to Entrez IDs
gene_ids <- bitr(all_sig_genes, 
                 fromType = "ENSEMBL", 
                 toType = "ENTREZID", 
                 OrgDb = org.Hs.eg.db)

mapped_genes <- unique(gene_ids$ENTREZID)

# Step 4: GO enrichment
ego_all <- enrichGO(
  gene         = mapped_genes,
  OrgDb        = org.Hs.eg.db,
  keyType      = "ENTREZID",
  ont          = "BP",
  pvalueCutoff = 0.05,
  readable     = TRUE
)

# Step 5: KEGG enrichment
ekegg_all <- enrichKEGG(
  gene         = mapped_genes,
  organism     = 'hsa',
  pvalueCutoff = 0.05
)

# Step 6: Plot
barplot(ego_all, showCategory = 10, title = "GO Enrichment: All Bacterial Agents")
barplot(ekegg_all, showCategory = 10, title = "KEGG Enrichment: All Bacterial Agents")

```
#All

```{r}
# Load required packages
library(DESeq2)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(ggplot2)
library(dplyr)

# List of all agents compared to control
all_agents <- c("IL-1B", "Bacteroides_dorei_BD", "Escherichia_coli_EC", "Ruminococcus_gnavus_RG")
reference <- "Serum-Free_Control"

# Create output folders (optional)
dir.create("enrichment_plots", showWarnings = FALSE)

for (agent in all_agents) {
  
  # Step 1: Run DESeq2 results for each agent vs control
  res <- results(dds, contrast = c("agent", agent, reference))
  res <- na.omit(res)
  res$gene <- rownames(res)

  # Step 2: Filter significant genes
  sig_genes <- res %>%
    as.data.frame() %>%
    filter(padj < 0.05 & abs(log2FoldChange) > 1) %>%
    pull(gene)

  # Step 3: Convert ENSEMBL to Entrez
  gene_ids <- bitr(sig_genes, fromType = "ENSEMBL", 
                   toType = "ENTREZID", 
                   OrgDb = org.Hs.eg.db)
  mapped_genes <- gene_ids$ENTREZID

  # Step 4: GO enrichment (Biological Process)
  ego <- enrichGO(
    gene         = mapped_genes,
    OrgDb        = org.Hs.eg.db,
    keyType      = "ENTREZID",
    ont          = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05,
    readable      = TRUE
  )

  # Step 5: KEGG enrichment
  ekegg <- enrichKEGG(
    gene         = mapped_genes,
    organism     = 'hsa',
    pvalueCutoff = 0.05
  )
  print(paste0("Saving plot to: ", getwd(), "/enrichment_plots/GO_", agent, ".png"))


  # Step 6: Save GO plot
  png(filename = paste0("enrichment_plots/GO_", agent, ".png"), width = 1000, height = 800)
  barplot(ego, showCategory = 10, title = "GO Enrichment")
  dev.off()

  # Step 7: Save KEGG plot
  png(filename = paste0("enrichment_plots/KEGG_", agent, ".png"), width = 1000, height = 800)
  barplot(ekegg, showCategory = 10, title = "KEGG Enrichment")
  dev.off()

  cat("✅ Completed:", agent, "\n\n")
}

```

